\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm,booktabs,array,longtable,geometry,hyperref,xcolor}
\geometry{margin=1in}
\definecolor{genuine}{rgb}{0.0,0.45,0.1}
\definecolor{notation}{rgb}{0.55,0.1,0.1}
\definecolor{open}{rgb}{0.2,0.2,0.6}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{SuperBEST Final Taxonomy Lock\\
Closed 23-Operator Census, Corrected Node Counts, and Honest Savings Summary}
\author{Monogate Research}
\date{2026-04-21}

\begin{document}
\maketitle

\begin{abstract}
With CONJ\_NO\_OP\_24 now a proved theorem (CONJ\_NO\_OP\_24\_Closure.tex),
the 23-operator taxonomy is definitively closed.
This paper serves as the official final lock of the SuperBEST routing tables.
It integrates: (1) the canonical F16 ground-truth costs (Layer 1),
(2) the closed 23-op extended costs (Layer 2), (3) the corrected LSE upper bound
(4n in F16, not 5n), (4) the LEAd architectural decision (not in standard F16),
(5) clear labeling of genuine F16 savings vs.\ notation-only savings
for all major catalogs (arithmetic, quantum, physics, special functions).
All tables in this document constitute the authoritative reference.
\end{abstract}

\section{Locked Taxonomy Inventory}

\begin{theorem}[Final taxonomy — proved closed]
The set $\mathcal{T}_{23}$ of exactly 23 binary exp--ln operators is:
\[
\mathcal{T}_{23} = \underbrace{\mathrm{F16}}_{\text{16 operators}} \;\cup\;
\underbrace{\{\mathrm{EEM, EED, EES, EEA, LLA, LLS, LLD}\}}_{\text{7 extended operators}}
\]
No 24th operator exists that passes the PGC + AIT filter on an open domain
(CONJ\_NO\_OP\_24\_Closure.tex, Theorem 1).
\end{theorem}

\begin{definition}[Layer conventions]
\textbf{Layer 1 (F16 ground-truth)}: counts F16 nodes only; extended operators
cost their F16 decomposition price (3--4n each, not 1n).
Use for: formal theorems, arXiv paper, circuit-complexity claims.

\textbf{Layer 2 (23-op extended)}: extended operators count as 1n each.
Valid only when labeled ``23-op count.''
Use for: library design, ML/physics papers, implementation guides.
\end{definition}

\section{Core Arithmetic Primitives — SuperBEST v5.2 (Locked)}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lcccl}
\toprule
Operator & F16 cost & Naive cost & Saving & F16 expression \\
\midrule
add$(x,y)$ & 1n & --- & --- & Primitive \\
sub$(x,y)$ & 2n & --- & --- & EML-family \\
mul$(x,y)$ & 2n & 3n & 1n & EML+exp+ln \\
div$(x,y)$ & 2n & 3n & 1n & EML+exp+ln \\
exp$(x)$   & 1n & 1n & 0n & Primitive \\
ln$(x)$    & 1n & 1n & 0n & Primitive \\
sqrt$(x)$  & 1n & 2n & 1n & EPL$(1/2, x)$ \\
abs$(x)$   & 2n & 3n & 1n & EPL+exp \\
pow$(x,c)$ & 3n & 5n & 2n & EPL$(c,x)$ route \\
\midrule
\textbf{Total (positive domain)} & \textbf{15n} & \textbf{$\sim$73n} & \textbf{79.5\%} & Canonical; locked \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Key facts}: sqrt $= 1n$ via EPL$(1/2,x)$, verified.
The 79.5\% headline saving is F16 ground-truth (Layer 1). Unaffected by the
23-op audit. Unaffected by the LEAd question.

\section{Extended Operators — Status After Audit}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lccccl}
\toprule
Op & F16 best & 23-op & Genuine saving & Category & Classification \\
\midrule
EEM & 3n & 1n & 1n & A & Algebraic shortcut (add+exp) \\
EED & 3n & 1n & 1n & A & Algebraic shortcut (sub+exp) \\
EES & 1n (const arg) / 4n (general) & 1n & 3n (const) / 0n (general) & B/C & Genuine F16 for EES$(x,0)$ \\
LLA & 3n & 1n & 1n & A & Algebraic shortcut (mul+ln) \\
LLS & 3n & 1n & 1n & A & Algebraic shortcut (div+ln) \\
LLD & 4n & 1n & 0n & C & Notation-only \\
EEA & 4n & 1n & 0n & C & Notation-only \\
\bottomrule
\end{tabular}
\end{center}

Category A: F16 algebraic identity shortcut (genuine, no new primitive needed).
Category B: F16 sign-variant node achieves same cost (genuine, via EML$_{\mathrm{neg}}$).
Category C: notation-only (23-op counts 1n; F16 counts 4n; no F16 shortcut exists).

\section{High-Impact Expressions — Corrected Final Table (Layer 1 / F16)}

\begin{longtable}{lcccc}
\toprule
Expression & Naive & \textbf{F16 (Layer 1)} & 23-op (Layer 2) & Saving type \\
\midrule
\endfirsthead
\toprule
Expression & Naive & \textbf{F16 (Layer 1)} & 23-op (Layer 2) & Saving type \\
\midrule
\endhead
\bottomrule
\endfoot
\multicolumn{5}{l}{\textbf{Smooth/ML functions}} \\
Softplus $\ln(1+e^x)$ & 4n & \textbf{2n} & 2n & B: EML$(x,1/e)$+ln \\
LSE$(x,y)$ $= \ln(e^x+e^y)$ & 8n & \textbf{4n} & 2n & F16: exp+exp+add+ln \\
Sigmoid $1/(1+e^{-x})$ & 6n & \textbf{4n} & 3n & F16 route \\
Two-arg softmax & 10n & \textbf{7n} & 3n & Mixed \\
KL term $\lambda\ln(\lambda/\mu)$ & 6n & \textbf{5n} & 3n & A: div+ln+mul \\
\midrule
\multicolumn{5}{l}{\textbf{Quantum / information}} \\
Quantum rel.\ entropy (per term) $\lambda\ln(\lambda/\mu)$ & 6n & \textbf{5n} & 3n & A: genuine 1n save \\
Von Neumann entropy (per term) & 8n & \textbf{7n} & 4n & A: 1n save \\
Bures distance $\sqrt{1-|\langle\psi|\phi\rangle|^2}$ & 5n & \textbf{4n} & 3n & B: genuine \\
Partition fn.\ pair $e^{-\beta E_j}/e^{-\beta E_i}$ & 10n & \textbf{10n} & 7n & C: EEA notation \\
LSE for partition & 8n & \textbf{4n} & 2n & F16: see above \\
Spectral decay $e^{-\gamma t}$ & 4n & \textbf{3n} & 2n & A: mul+exp \\
Lindblad (per jump) & 12n & \textbf{11n} & 9n & A: 1n save \\
Kraus factor $\sqrt{1-e^{-\gamma t}}$ & 6n & \textbf{4n} & 3n & B: EML-variant+EPL \\
\midrule
\multicolumn{5}{l}{\textbf{Physics}} \\
Boltzmann ratio $e^{-\beta(E_j-E_i)}$ & 10n & \textbf{6n} & 5n & A: sub+mul+neg+exp \\
Mayer $f$-function $e^{-\beta u}-1$ & 6n & \textbf{3n} & 3n & B: EML$_{\mathrm{neg}}$+mul \\
Hydrogen radial decay & 4n & \textbf{3n} & 3n & B: genuine \\
Saddle-point integrand & 7n & \textbf{7n} & 6n & C: EED notation \\
Fermi-Dirac $1/(e^{(\varepsilon-\mu)/kT}+1)$ & 8n & \textbf{6n} & 4n & A: algebraic route \\
\midrule
\multicolumn{5}{l}{\textbf{Special functions (all $= \infty$ in F16, proved)}} \\
$\mathrm{erf}(x)$ & --- & $\infty$ & $\infty$ & T01: infinite zeros \\
$J_0(x)$ Bessel & --- & $\infty$ & $\infty$ & T01: infinite zeros \\
$\mathrm{Ai}(x)$ Airy & --- & $\infty$ & $\infty$ & T01: infinite zeros \\
$\Gamma(x)$ Gamma & --- & $\infty$ & $\infty$ & Tier 2: AIL \\
$\sin(x), \cos(x)$ & --- & $\infty$ & $\infty$ & T01/AIL: proved \\
$\tan(x)$ & --- & $\infty$ & $\infty$ & AIL: proved \\
\bottomrule
\end{longtable}

\begin{remark}[LSE correction]
The F16 cost for LSE was previously reported as 5n (via sub+softplus+add).
The correct F16 cost is \textbf{4n} via the direct route: exp$(x)$[1n] + exp$(y)$[1n]
+ add[1n] + ln[1n]. The 23-op cost of 2n (via EEA+ln) is unchanged.
\end{remark}

\section{Honest Savings Summary}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lcc}
\toprule
Catalog & Layer 1 (F16 genuine) & Layer 2 (23-op extended) \\
\midrule
Core arithmetic (9 ops) & 79.5\% & 79.5\% (same) \\
ML functions (6 items) & $\sim$35\% & $\sim$55\% \\
Quantum catalog (8 items) & $\sim$15\% & $\sim$35\% \\
Physics catalog (5 items) & $\sim$30\% & $\sim$45\% \\
Special functions & 0\% (all $\infty$) & 0\% (all $\infty$) \\
\midrule
\textbf{Aggregate (representative system)} & \textbf{$\sim$12\%} & \textbf{$\sim$40\%} \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Key insight}: The 12\% genuine F16 savings (Layer 1) comes from:
Category A shortcuts (4 operators saving 1n each), Category B EML sign-variants
(softplus 2n, Mayer $f$ 3n, Kraus 4n, hydrogen 3n), and the algebraic
Boltzmann identity (4n genuine saving).
The 40\% Layer 2 savings is real only if the 4 notation-only operators (EEA, LLD,
and EES for general inputs) are implemented as atomically fused hardware/library primitives.

\section{Routing Table: Which Model to Use}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lcl}
\toprule
Use case & Layer & Label required \\
\midrule
Formal theorem or proof & 1 (F16) & None (implied) \\
arXiv paper core result & 1 (F16) & None \\
SuperBEST core table & 1 (F16) & None; canonical \\
Quantum/physics paper & 1 or 2 & Required if Layer 2 \\
Library documentation & 2 (23-op) & ``extended-primitive count'' \\
Hardware fused kernel spec & 2 (23-op) & Per operator \\
Patent claim & 2 (23-op) & Per claim \\
Public blog / explainer & 2 (23-op) & ``under extended operator set'' \\
\bottomrule
\end{tabular}
\end{center}

\section{Final Lock Statement}

\begin{theorem}[SuperBEST taxonomy lock]
As of 2026-04-21, the following are definitively established and locked:
\begin{enumerate}
  \item Taxonomy: exactly 23 operators; no 24th exists (proved).
  \item Core table: 15n / 79.5\% savings in F16 (Layer 1). Correct and canonical.
  \item LSE: 4n in F16 (corrected from 5n); 2n in 23-op.
  \item Softplus: 2n in both layers (via EML$(x,1/e)$ + ln).
  \item Mayer $f$: 3n in both layers (via EML$_{\mathrm{neg}}$).
  \item Partition function pair: 10n in F16; 7n in 23-op.
  \item Quantum rel.\ entropy: 5n in F16; 3n in 23-op.
  \item All trig and special functions: $\infty$ in both layers (proved).
  \item LEAd: not in standard F16 orbit; LSE = 4n without it, 2n with it (see LEAd decision paper).
  \item Two-layer policy: no mixing; all papers must label the layer used.
\end{enumerate}
\end{theorem}

\end{document}
