% RH-EML Forward Direction: Honest Analysis
% Session S112 -- monogate private research document
% Status: CONJECTURE with partial evidence. NOT a proof of RH.

\section*{RH-EML: Honest Status of the Forward Direction}

\subsection*{The Conjecture}

The RH-EML conjecture (Sessions 48--49) states:
\begin{quote}
The Riemann Hypothesis is \emph{equivalent} to the claim that the function
$f_\sigma(t) = \mathrm{Re}(\zeta(\sigma + it))$ is EML-$\infty$ (as a
function of $t$) if and only if $\sigma = 1/2$.
\end{quote}

\subsection*{What Is Actually Proved}

\textbf{Forward direction (conditional):}
Assuming the Riemann Hypothesis holds, the non-trivial zeros of $\zeta$
lie on $\sigma = 1/2$. Hardy (1914) proved infinitely many zeros lie
on the critical line. Together, these imply $f_{1/2}(t) = \mathrm{Re}(\zeta(1/2+it))$
has infinitely many zeros (one at each ordinate of a zero on the critical line).
By the Infinite Zeros Barrier (T01), $f_{1/2} \in \mathrm{EML}\text{-}\infty$.

This is trivially true and says nothing new about RH.

\textbf{Why the converse is not a proof of RH:}
The converse would say: if $f_\sigma \in \mathrm{EML}\text{-}\infty$, then
$\sigma = 1/2$.  This requires showing that $f_\sigma$ has infinitely many
zeros \emph{only for $\sigma = 1/2$}.  But this is exactly a restatement
of RH (zeros of $\zeta$ are on $\sigma=1/2$) --- not an independent argument.

The EML depth classification adds no information to the RH question.

\subsection*{What the Numerical Evidence Shows}

From Sessions 48--49: for 20 numerically computed Riemann zeros (all on $\sigma=1/2$),
$f_{1/2}(t)$ changes sign near each ordinate.  For $\sigma = 0.7$, fewer
sign changes are observed in the same range.

This is \emph{consistent with} RH (and with Conrey's 2/5 result, and with
$10^{13}$ verified zeros) but is not a proof.

\subsection*{Verdict}

The RH-EML connection is:
\begin{itemize}
  \item A \textbf{useful framing}: RH is equivalent to $\zeta$ being EML-$\infty$
    on the critical line in a specific technical sense.
  \item \textbf{Not a proof path}: EML depth analysis cannot distinguish
    ``infinitely many zeros because RH holds'' from ``infinitely many zeros
    by some other reason.''
  \item \textbf{Not to be presented publicly} as a proof or near-proof of RH.
\end{itemize}

The honest public statement is: ``The EML Infinite Zeros Barrier provides a
new language for the zero-count structure of $\zeta$, but does not resolve RH.''

\subsection*{What Would Be Needed}

A genuine RH proof via EML would require one of:
\begin{enumerate}
  \item Showing that the EML depth of $f_\sigma$ is a \emph{monotone function}
    of $|\sigma - 1/2|$ --- but this would need RH to establish.
  \item Finding an EML-based functional equation or symmetry argument that
    forces zeros to the critical line --- no such argument is known.
  \item Using the EML-Fourier decomposition to detect off-critical zeros ---
    speculative; would require understanding the full zero distribution.
\end{enumerate}

None of these are available.  RH remains open.

\subsection*{Sessions to Deprioritise}

Sessions 48--49, 89--90, and the RH-related content in sessions 100+ are
\emph{archived as speculative} per the session classification rules.
They should not appear in Layer 1 (public repo) or the arXiv preprint
without the above disclaimer.
